Reducing Acceptance Marks in Emerson-Lei Automata by QBF Solving

### 2023-06-01 paper/main.tex

Recall that we aim to reduce the number of acceptance marks in a given
TELA $\mA =(Q,M,\Sigma,\delta,q_I,\varphi)$ without altering its
structure and language. In other words, we look for
\begin{itemize}
\item a set $\oM$ of acceptance marks satisfying $|\oM|<|M|$,
\item an acceptance formula $\psi$ over $\oM$, and 
\item a function
  $\rem:\delta\rightarrow Q \times \Sigma \times 2^{\oM}\times Q$
  assigning new marks to transitions (i.e., for each
  $t=(p,a,M',q)\in\delta$, we assume that $\rem(t)=(p,a,\oM',q)$ for
  some $\oM'\subseteq\oM$)
  %
  such that the automaton $\mB=(Q,\oM,\Sigma,\rem(\delta),q_I,\psi)$
  is equivalent to $\mA$.
\end{itemize}
% To simplify the task, 
% To ensure that the modified automaton is equivalent to the original one,
We will actually look for $\psi$ and $\rem$ such that each run
$\pi=t_0t_1t_2\ldots$ of $\mA$ is accepting if and only if the run
$\rem(t_0)\rem(t_1)\rem(t_2)\ldots$ of $\mB$ is accepting. This
requirement clearly guarantees the equiavalence of $\mA$ and $\mB$. We
note that preserving the acceptance of individual runs is not a
necessary condition for language equivalence of $\mA$ and $\mB$ and
there exist automata where relaxing this requirement can lead to
a bigger reduction of acceptance marks (see
Figure~\ref{fig:counter}). However, looking for $\psi$ and $\rem$
preserving the acceptance of individual runs makes the problem easier
as we can, for example, ignore the labelling of transitions by the
elements of $\Sigma$.
\begin{figure}[t]
  \centering
  \begin{tikzpicture}[smallautomaton]
    \begin{scope}[initial angle=180,xshift=-3.5cm]
      \node[state,initial] (p) at (0,0) {};
      \node[state] (q) at (2,0) {};
      \path[->,auto]
      (p) edge[out=120,in=60,looseness=10] pic {acc={0}{0}} pic {l=$a$} (p)
      (p) edge[out=-60,in=-120,looseness=10] pic {acc={1}{1}} pic {l=$b$} (p)
      (p) edge[out=30,in=150] pic {l=$a$} (q)
      (q) edge[out=-150,in=-30] pic {l=$b$} (p);

@@

  \caption{Both automata accept the same language of words
    $w\in\{a,b\}^\omega$ that contain infinitely many occurences of
    both $a$ and $b$. Each accepting run of the left automaton has to
    contain infinitely many occurrences of both transitions looping on
    the initial state. Hence, there does not exist any automaton with
    the same accepting runs as the left automaton and less than two
    accetpance marks.}
  \label{fig:counter}
\end{figure}

\fixme{TADY TO HONZA JESTE UPRAVUJE}
We start with several simple observations. Acceptance of a run $\pi$
is fully determined by the set $\minf(\pi)$.
% of acceptance marks that
% appear infinitely often on $\pi$, which is in turn fully determined by
% the set $\minf(\pi)$ of transitions appearing infinitely often in
% $\pi$.
\begin{lemma}
  A set $T\subseteq\delta$ is a cycle if and only if
  there is a run $\pi$ such that $T=\minf(\pi)$.
\end{lemma}

As $\minf(\pi)$ is the set of transitions that appear infinitely many
times in $\pi$, there has to be a suffix $\pi'$ of $\pi$ containing
only transitions of $\minf(\pi)$. Let $p$ be the first state of
$\pi'$. As each transition of $\pi'$ appears infinitely many times in
$\pi$ and thus also in $\pi'$, there has to be a finite prefix $\rho$
of $\pi'$ such that $\rho$ is a path from $p$ to $p$ that contains all
transitions of $\minf(\pi)$. In other words, the set $\minf(\pi)$ of a
run $\pi$ is always a cycle. Moreover, as we assume that 
%\footnote{We can also prove that for each cycle $T$ there exists a run $\pi$ such that $T=\minf(\pi)$.}

\fixme{ODTUD DAL JE TO PUVODNI, ale taky se to trochu bude menit!}

Further, each set $\minf(\pi)$ for a run $\pi$ is a cycle and vice
versa. Hence, our reduction method is looking for a new acceptance
formula $\psi$ and function $\rem$ such that for each cycle
$T\subseteq\delta$, it holds that $T$ satisfies $\varphi$ if and only
if $\rem(T)$ satisfies $\psi$. This can be roughly denoted by the
formula
\[
  \forall T\subseteq\delta~.~\mcycle(T)\implies\big(\msat_\varphi(T) ~\iff~ \msat_{\psi}(\rem(T))\big).
\]
In fact, this corresponds to the shape of the QBF we will
construct. As we are looking for $\psi$ and $\rem$ such that the
formula holds, the subformula $\msat_{\psi}(\rem(T))$ contains many
free variables representing possible instances of $\psi$ and
$\rem$. If the formula is satisfiable, then each of its models encodes
a desired instance of $\psi$ and $\rem$. In the following, we assume
that we are looking for a new acceptance formula $\psi$ in DNF. The
choice of DNF is not fundamental, but inherited from our previous
attempt to reduce acceptance formulas. The presented method can be
easily adapted to look for $\psi$ in \emph{conjunctive normal form
  (CNF)} or in a different shape.

>>>>

Recall that we aim to reduce the number of acceptance marks in a given
TELA $\mA =(Q,M,\Sigma,\delta,q_I,\varphi)$ without altering its
structure and language. In other words, we look for
\begin{itemize}
\item a set $\oM$ of acceptance marks satisfying $|\oM|<|M|$,
\item an acceptance formula $\psi$ over $\oM$, and 
\item a function
  $\rem:\delta\rightarrow Q \times \Sigma \times 2^{\oM}\times Q$
  assigning new marks to transitions (i.e., for each
  $t=(p,a,M',q)\in\delta$, we assume that $\rem(t)=(p,a,\oM',q)$ for
  some $\oM'\subseteq\oM$)
  %
  such that the automaton $\mB=(Q,\oM,\Sigma,\rem(\delta),q_I,\psi)$
  is equivalent to $\mA$.
\end{itemize}
% To simplify the task, 
% To ensure that the modified automaton is equivalent to the original one,
We will actually look for $\psi$ and $\rem$ such that each run
$\pi=t_0t_1t_2\ldots$ of $\mA$ is accepting if and only if the run
$\rem(t_0)\rem(t_1)\rem(t_2)\ldots$ of $\mB$ is accepting. This
requirement clearly guarantees the equiavalence of $\mA$ and $\mB$,
but it is not a necessary condition for the equivalence. Indeed, there
exist automata where relaxing this requirement can lead to a bigger
reduction of acceptance marks (see Figure~\ref{fig:counter}). However,
looking for $\psi$ and $\rem$ that preserve the acceptance of individual
runs makes the problem easier as we can, for example, ignore the
labelling of transitions by the elements of $\Sigma$.
\begin{figure}[t]
  \centering
  \begin{tikzpicture}[smallautomaton]
    \begin{scope}[initial angle=180,xshift=-3.5cm]
      \node[state,initial] (p) at (0,0) {};
      \node[state] (q) at (2,0) {};
      \path[->,auto]
      (p) edge[out=120,in=60,looseness=10] pic {acc={0}{0}} pic {l=$a$} (p)
      (p) edge[out=-60,in=-120,looseness=10] pic {acc={1}{1}} pic {l=$b$} (p)
      (p) edge[out=30,in=150] pic {l=$a$} (q)
      (q) edge[out=-150,in=-30] pic {l=$b$} (p);

@@

  \caption{The left automaton accepts the words % $w\in\{a,b\}^\omega$
    that contain infinitely many occurences of both $a$ and $b$. Each
    accepting run of the left automaton has to contain infinitely many
    occurrences of both transitions looping on the initial
    state. Hence, there does not exist any automaton with the same
    accepting runs as the left automaton and less than two accetpance
    marks. The right automaton accepts the same language using one
    acceptance mark and a different set of accepting runs.}
  \label{fig:counter}
\end{figure}

Our reduction method is based on two facts. First, the acceptance of a
run $\pi$ is fully determined by $\minf(\pi)$. Second, each set
$\minf(\pi)$ is a cycle and vice versa.

% Before we We start with a reformulation of our requirement on $\psi$
% and $\rem$.  Recall that a run $\pi$ of $\mA$ is accepting if and
% only iff $\minf(\pi)$ satisfies $\varphi$.  Further, the following
% lemma says that each set $\minf(\pi)$ is a cycle and vice versa.

\begin{lemma}
  A set $T\subseteq\delta$ is a cycle if and only if
  there is a run $\pi$ such that $\minf(\pi)=T$.
\end{lemma}
\begin{proof}
  To prove the direction `$\Longrightarrow$', we assume that $T$ is a
  cycle. The definition says that there exists a path $\tau$ from a
  state $p$ to the same state containing each transition of $T$ at
  least once and no transition outside $T$. As our automata contain
  only reachable states, there exists a path $\rho$ from the initial
  state $q_I$ to $p$ or $p=q_I$ and we set $\rho=\varepsilon$. The
  infinite sequence $\pi=\rho.\tau^\omega$ is a run satisfying
  $\minf(\pi)=T$.

  To prove the direction `$\Longleftarrow$', we consider a run
  $\pi$.  As $\minf(\pi)$ is the set of transitions that appear
  infinitely many times in $\pi$, there has to be a suffix $\pi'$ of
  $\pi$ containing only transitions of $\minf(\pi)$. Let $p$ be the
  first state of $\pi'$. As each transition of $\pi'$ appears
  infinitely many times in $\pi$ and thus also in $\pi'$, there has to
  be a finite prefix $\rho$ of $\pi'$ such that $\rho$ is a path from
  $p$ to $p$ that contains all transitions of $\minf(\pi)$. In other
  words, the set $\minf(\pi)$ is a cycle.
\end{proof}

Hence, our goal can be reformulated as follows. We look for a new
acceptance formula $\psi$ and a function $\rem$ such that for each
cycle $T\subseteq\delta$, it holds that $T$ satisfies $\varphi$ if and
only if $\rem(T)$ satisfies $\psi$. This can be roughly denoted by the
formula
\[
  \forall T\subseteq\delta~.~\mcycle(T)\implies\big(\msat_\varphi(T) ~\iff~ \msat_{\psi}(\rem(T))\big).
\]
In fact, this corresponds to the shape of the QBF we will
construct. As we are looking for $\psi$ and $\rem$ such that the
formula holds, the subformula $\msat_{\psi}(\rem(T))$ contains many
free variables representing possible instances of $\psi$ and
$\rem$. If the formula is satisfiable, then each of its models encodes
a desired instance of $\psi$ and $\rem$. In the following, we assume
that we are looking for a new acceptance formula $\psi$ in DNF. The
choice of DNF is not fundamental, but inherited from our previous
attempt to reduce acceptance formulas. The presented method can be
easily adapted to look for $\psi$ in \emph{conjunctive normal form
  (CNF)} or in a different shape.
